Software variability in fMRI analysis:\\ comparing between-tool and numerical errors

### 2021-12-03 paper/paper.tex
New introduction

\title{Comparing tool variability and numerical variability in fMRI analyses}

@@

Numerical and software variability broadly affect structural, diffusion, and functional MRI analyses. Numerical
variability originates in software updates or code
parallelization, whereas software variability reflects discrepancies between
models implemented in different analysis software packages. While numerical
and software variability were both shown to impact analysis outcomes, the
extent to which these sources of variability compare for a given
analysis remains understudied. This work presents a comparison of
numerical and software variability for group-level and subject-level functional MRI analyses.
We reproduced a previous comparison between functional MRI analysis
software packages FSL, AFNI, and SPM, which we extended to measure
numerical variability through Monte-Carlo arithmetic.
Although We found that between-tool variability was larger than numerical variability,
we identified brain regions that numerical perturbations simulated between-tool variability.
Furthermore, we found more numerical instability in thresholded maps than unthresholded ones
and individual analysis than group-level analysis results.

@@

% Data analysis workflows in many scientific domains have become increasingly complex and flexible.
Recent studies highlighted the instability of the neuroimaging pipelines depending on the computing platform,
software package, and even tool versions. Changes in the computational
environment such as compilers, libraries, operating systems may introduce small numerical errors and create
significantly different results in unstable pipelines~\cite{Glatard2015,Gronenschild2012,salari2020spot}.
Moreover, the impact of methodological changes on fMRI analyses has been investigated extensively~\cite{bowring2019exploring,botvinik2020variability,bhagwat2021understanding,carp2012plurality}.
For instance, in related works, it has been shown that running the same fMRI experiments by different teams can substantially affect
scientific conclusions~\cite{botvinik2020variability,carp2012plurality};
replication of fMRI experiments using the three most well-known software packages can influence the final determining areas of
brain activations~\cite{bowring2019exploring}; %bowring2021isolating.
also, the choice of preprocessing pipelines on neuroimaging cortical surface analyses is compound with instabilities~\cite{bhagwat2021understanding}.
% In the presence of such instabilities, it is often hard to trust the data processing results. % validity of the computational results.

In such a heterogeneous environment, numerical instability is an essential issue for reproducibility.
Numerical instability is a characteristic of pipelines that results from the influence of the floating-point arithmetics
and iterative convergence of numerical errors~\cite{freitas2002issue}.
Stochastic arithmetic approaches such as Monte-Carlo arithmetic (MCA)~\cite{Parker1997-qq} have been used to study the impact of numerical errors
originating in floating-point computations in mathematical libraries.
In~\cite{salari2021accurate}, we quantified the numerical stability of the HCP preprocessing pipeline~\cite{glasser2013} based on the MCA method
by creating a Fuzzy environment, so that instrument mathematical functions are implemented in mathematical system libraries.
As a result of numerical perturbations, we discovered a very low numerical precision in the result images comparable to the operating system variability.
In a related study~\cite{kiar2020numerical}, the instability of results was explored by instrumenting a connectome estimation pipeline with the MCA technique.
These works demonstrate the necessity of numerical uncertainty quantification for understanding related issues that hamper the computational reproducibility of analyses.

\tristan{We investigate two questions: (1) how do numerical and between-tool variability compare? (2) are they correlated?}

In this work, we reproduce an fMRI analysis in~\cite{bowring2019exploring} with different neuroimaging software packages in the presence of the Fuzzy environment,
and then quantify the variability across tools and the numerical variability in results.
We call the cross-software variation and numerical variation, between tool (BT) and within tool (WT) variability, respectively.
In fact, the WT variability shows results across software packages in the Fuzzy environment.
The primary objective of this study is to answer these two questions: 1) how the fMRI analyses across tools are numerically stable?
2) how the numerical variability is in comparison with the tool variability?
This comparative study reveals the importance of numerical variability and motivates research studies to evaluate the numerical uncertainty of the pipelines.

% We start to reproduce an analysis from Bowring, this can be as a practice for reproducibility manner in the community.
% In particular, we investigate the effect of 1) between software 2) within software (numerical)
% We present a comparative assessment of group-level analysis of an fMRI pipeline.

>>>>

\title{Software variability in fMRI analysis:\\ comparing between-tool and numerical errors}

@@

Numerical and software variability broadly affect structural, diffusion, and functional MRI analyses. Numerical
variability originates in software updates or code
parallelization, whereas software variability reflects discrepancies between
models implemented in different analysis software packages. While numerical
and software variability were both shown to impact analysis outcomes, the
extent to which these sources of variability compare for a given
analysis remains understudied. This work presents a comparison of
numerical and software variability for group-level and subject-level functional MRI analyses.
We reproduced a previous comparison between functional MRI analysis
software packages FSL, AFNI, and SPM, which we extended to measure
numerical variability through Monte-Carlo arithmetic.
Although We found that between-tool variability was larger than numerical variability,
we identified brain regions that numerical perturbations simulated between-tool variability.
Furthermore, we found more numerical instability in thresholded maps than unthresholded ones
and individual analysis than group-level analysis results. \tristan{Abstract to be fixed}

@@

Software variability has been highlighted as an important source of error
in fMRI analysis, originating from mainly two factors: Between-tool
variability denotes differences observed in results produced by different
software toolboxes --- such as FSL, AFNI and SPM --- when analyzing the same dataset
using similar methods. Numerical variability, in contrast, denotes
differences observed in results produced by the same analytical toolbox
executed in different environments – such as different operating systems or
different hardware platforms. This manuscript investigates the extent to
which both types of variability relate to each other.

Numerical variability stems from errors introduced in floating point
computations by numerical rounding, cancellation and absorption~\cite{muller2018handbook}. In
practice, the occurrence of such errors is affected by variations in
hardware, software dependencies, compilation, parallelization, and most
likely other factors. In general, the magnitude of such errors and their
variations is in the range of 1 unit in the last place (ulp) --- also known
as machine error, the difference between two successive floating-point
numbers. However, when computations depend on imprecise libraries,
numerical error may vary by several ulps. For instance, the numerical error
of the complementary error function (erfc) is larger than 3 ulps in several
mathematical libraries~\cite{zimmermann:hal-03141101}, implying that an update of this function could introduce
numerical perturbations larger than 3 ulps. Due to numerical instability,
machine error may amplify during its propagation through the analysis,
resulting in considerable loss of precision in final results. In fMRI,
numerical variability produced by Linux updates was shown to result in Dice
coefficients ranging from 0.0 to 1.0 for ICA components produced by FSL
MELODIC~\cite{Glatard2015}.

Between-tool variability is not a surprising concept for software
developers: different programs executed on identical data may obviously
produce different results. However, in fMRI, analysis programs all aim to
estimate brain activity from measured variations in the BOLD signal, which
is not expected to be measurably impacted by computational artifacts.
Therefore, the observation reported in~\cite{bowring2019exploring} that the three
main fMRI analysis toolboxes may result in marked
analytical differences, such as Dice coefficients ranging from 0.000 to
0.684 in comparisons of thresholded group-level statistic maps, has raised
substantial concern. In a follow-up study~\cite{bowring2021isolating}, the same
authors isolate the analytical components where the pipelines diverge,
highlighting a major impact of the signal model and suggesting that the
root-cause for between-tool variability depends on the analysis design and
task paradigm.  \tristan{[refer to Xinhui's pre-print].}

We investigate the relationship between numerical and between-tool
variability through two main questions: (1) how do both types of
variability compare in magnitude, and (2) is there an association between
them. The first question determines the extent to which each variability
source may impact current fMRI results, and which one should be addressed
in priority. The second question may provide deeper insights on the origins
of software variability. Conceptually, both types of variability may
reflect a single solution space shaped by local minima and
ill-conditioning, which could be explored through numerical or software
perturbations.
